---
title: "EDA of Datasets"
header-includes:
  - \usepackage[dvipsnames]{xcolor}
  - \definecolor{darkblue}{gray}{0.45}
  - \definecolor{darkblue}{RGB}{25,25,112}
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \fancyhead[CO,CE]{\thepage}
  - \fancyfoot[CO,CE]{Alex Clark - Assignment \#3}
output: pdf_document
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)

c("data.table",
  "tidyverse",
  "plyr",
  "dtplyr",
  "TTR",
  "astsa",
  "lubridate",
  "tsbox",
  "tseries",
  "xts",
  "zoo",
  "forecast",
  "openxlsx",
  "mlbench",
  "caret",
  "random",
  "tsfeatures",
  "knitr"
  ) -> package_names  
for(package_name in package_names) {
  if(!is.element(package_name, installed.packages()[,1])) {
     install.packages(package_name,
                      repos = "http://cran.mtu.edu/")
  }
  library(package_name, character.only=TRUE,
          quietly=TRUE,verbose=FALSE)
}

rm(list=c("package_name", "package_names")) # clean up the environment
set.seed(61818)
options(show.signif.stars = FALSE)  # Don't confuse significance & effect size!
options(scipen=999) # Turn off scientifc notations
setwd("~/Documents/GitHub/GDAT626_Final")
```

```{r load in all dfs, include=FALSE}
c02 = read.csv("CO2.csv")
ice = read.csv("Ice.csv")
temp = read.csv("temp.csv")
```

### $CO_2$:
```{r head C02, echo=FALSE}
glimpse(c02)
```

### Arctic Sea Ice Level
```{r head ice, echo=FALSE}
glimpse(ice)
```

### Temperature:
```{r head temp, echo=FALSE}
glimpse(temp)
```

### Time Series of CO2:
```{r c02 time series, echo=FALSE}
c02.ts <- ts(c02$trend..season.corr., start = c(1958,3), frequency = 12)
head(c02.ts,20)
```

### Time Series of Ice Levels:
```{r ice time series, echo=FALSE}
ice.ts <- ts(ice$area, start = c(1979), end = c(2019), frequency = 1)
ice.ts
```

### Time Series of Temperature:
```{r temp time series, echo=FALSE}
temp.ts <- ts(temp$Lowess.5., start = c(1880), frequency = 1)
temp.ts
```

### Plot of $CO_2$:
```{r plot c02.ts, echo=FALSE}
plot(c02.ts, main = "CO2 Time Series", ylab = "Trend", xlab = "Year")
abline(reg=lm(c02.ts~time(c02.ts)), col = "blue")
plot(decompose(c02.ts))
boxplot(c02.ts~cycle(c02.ts), main = "CO2 Boxplot", ylab = "Trend", xlab = "Month")
```

### Plot of Ice Levels:
```{r plot ice.ts, echo=FALSE}
plot(ice.ts, main = "Time Series of Ice Levels", xlab = "Year", ylab = "Area")
abline(reg=lm(ice.ts~time(ice.ts)), col = "blue")
```

### Plot of Temperature:
```{r plot temp.ts, echo=FALSE}
plot(temp.ts, main = "Time Series of Temperature", xlab = "Year", ylab = "Lowess 5")
abline(reg=lm(temp.ts~time(temp.ts)), col = "blue")
```

### Hstogram and KS test of $CO_2$: 
```{r histogram of c02, echo=FALSE}
hist(c02.ts, main = "CO2 Histogram", xlab = "Trend")
ks.test(c02.ts, "dnorm")
```

### Hstogram and KS test of Ice Levels:
```{r histogram of ice, echo=FALSE}
hist(ice.ts, main = "Histogram of Ice Levels")
ks.test(ice.ts, "dnorm")
```

### Hstogram and KS test of Temperature:
```{r histogram of temp, echo=FALSE}
hist(temp.ts, main = "Histogram of Temperature")
ks.test(temp.ts, "dnorm")
```

### Test Stationary of $CO_2$: 
```{r stationary test of c02, echo=FALSE}
adf.test(c02.ts)
```

### Test Stationary of Ice Levels:
```{r stationary test of ice, echo=FALSE}
adf.test(ice.ts)
```

### Test Stationary of Temperature:
```{r stationary test of temp, echo=FALSE}
adf.test(temp.ts)
```

### Shift $CO_2$ by 5:
```{r shift c02 data, warning=FALSE, error=FALSE, echo=FALSE}
xn <- c02.ts
xnp1 <- shift(c02.ts,9)
plot(xnp1, xn, main = "Shift by 9 - CO2", xlab = "Shift", ylab = "CO2", type = "l") 
```

### Shift Ice Levels by 3:
```{r shift ice data, warning=FALSE, error=FALSE, echo=FALSE}
xn <- ice$area
xnp1 <- shift(ice$area, 3)
plot(xnp1, xn, main = "Shift by 3 - Ice", xlab = "Shift", ylab = "Ice", type = "l") 
```

### Shift Temperature by 5:
```{r shift temp data, warning=FALSE, error=FALSE, echo=FALSE}
xn <- temp$Lowess.5.
xnp1 <- shift(temp$Lowess.5., 5)
plot(xnp1, xn, main = "Shift by 5 - Ice", xlab = "Shift", ylab = "Lowess 5", type = "l") 
```

### ACF and PACF of $CO_2$:
```{r autocorrelate c02, echo=FALSE}
autoplot(acf(c02.ts, plot = FALSE)) + ggtitle("ACF - CO2") 
autoplot(pacf(c02.ts, plot = FALSE)) + ggtitle("PACF - CO2") 
```

### ACF and PACF of Ice Levels:
```{r autocorrelate ice, echo=FALSE}
autoplot(acf(ice.ts, plot = FALSE)) + ggtitle("ACF - Ice")
autoplot(pacf(ice.ts, plot = FALSE)) + ggtitle("PACF - Ice")
```

### ACF and PACF of Temperature:
```{r autocorrelate temp, echo=FALSE}
autoplot(acf(temp.ts, plot = FALSE)) + ggtitle("ACF - Temp")
autoplot(pacf(temp.ts, plot = FALSE)) + ggtitle("PACF - Temp")
```

### Periodogram of $CO_2$:
```{r periodogram c02, echo=FALSE}
TSA::periodogram(c02.ts, main = "CO2 Periodogram")
```

### Periodogram Ice Levels:
```{r periodogram ice, echo=FALSE}
TSA::periodogram(ice.ts, main = "Ice Periodogram")
```

### Periodogram Temperature:
```{r periodogram temp, echo=FALSE}
TSA::periodogram(temp.ts, main = "Temperature Periodogram")
```
